---
title: "class 10 - candy"
author: "jack olmstead"
format: gfm
---

```{r import}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy <- read.csv(url, row.names = 1)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

For some cursed reason, the tasteless, unwashed FifeThirtyEight analysts who put this frivilous exercise together didn't include Hot Tamales in the dataset. Probably because its win-rate was so game-breaking.
```{r}
candy["Haribo Sour Bears",]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library(skimr)
skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Win percent definite has a different scale than everything else.
```{r}
library(tidyverse)

candy$name <- rownames(candy)
neat <- gather(candy, "var", "value", -name)

ggplot(neat) +
  aes(x=name, y=value) +
  geom_col() +
  facet_wrap(~var)
```

> Q7. What do you think a zero and one represent for the candy$chocolate column?

  - 1: the candy is chocolate.
  - 0: the candy is not chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
ggplot(candy) +
  aes(x=winpercent) +
  geom_histogram()
```

> Q9. Is the distribution of winpercent values symmetrical?

Looks to be long-tailed on the right side.

> Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Definitely a chocolate bias
```{r}
ggplot(candy) +
  aes(x=winpercent, fill=factor(chocolate)) +
  geom_histogram(position="dodge2")
```

> Q.12. Is this difference statistically significant?

Very.
```{r}
t.test(
  candy[candy$chocolate == 1,]$winpercent, 
  candy[candy$chocolate == 0,]$winpercent
  )
```

> Q13. What are the five least liked candy types in this set?

```{r}
sorter <- sort(candy$winpercent, index.return=1, decreasing = T)
candy.sorted <- candy[sorter$ix,]
tail(candy.sorted)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy.sorted)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy.sorted) +
  aes(x=winpercent, y=rownames(candy.sorted)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy.sorted) +
  aes(x=winpercent, y=reorder(rownames(candy.sorted), winpercent)) +
  geom_col()
```

Let's make some useful color vectors!
```{r}
my.colors <- rep("black", nrow(candy))
my.colors[as.logical(candy$chocolate)] = "chocolate"
my.colors[as.logical(candy$bar)] = "brown"
my.colors[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy.sorted) +
  aes(x=winpercent, y=reorder(rownames(candy.sorted), winpercent)) +
  geom_col(fill=my.colors)
```

> Q17. What is the worst ranked chocolate candy?

Nik L Nip, whatever that is.

> Q18. What is the best ranked fruity candy?

Snickers, apparently? Although I don't think of this as a fruity candy.

```{r}
#| eval: false
library("ggrepel")

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my.colors) +
  geom_text_repel(col=my.colors, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy$bang.buck.ratio <- candy$winpercent / candy$pricepercent

ggplot(candy) +
  aes(x=bang.buck.ratio, y=reorder(rownames(candy), bang.buck.ratio)) +
  geom_col()
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy.sort.indicies.price <- sort(candy$pricepercent, decreasing = T, index.return=T)
head(candy[candy.sort.indicies.price$ix, ])
```


## exploring the dimensionality of these data
```{r}
library(corrplot)

cij <- cor(candy[, -13])
corrplot(cij)
```

```{r PCA}
pcs <- prcomp(candy[,-13], scale=T)
summary(pcs)
```

Plot PCs
```{r}
plot(pcs$x[,1], pcs$x[,2], col=my.colors, pch=16)
```

Add these PC coordinates to our candy data.
```{r}
candy <- cbind(candy, pcs$x[, 1:3])
```

```{r}
p <- ggplot(candy) +
  aes(x=PC1, y=PC2, 
      size=winpercent/100,
      color=my.colors) +
  geom_point()
p
```

Last, let's look at the loadings of our stuffs.
```{r}
par(mar=c(8,4,2,2))

barplot(pcs$rotation[,1], las=2, ylab="PC1 Contribution")
barplot(pcs$rotation[,2], las=2, ylab="PC2 Contribution")

# library(patchwork)
# p1+p2
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, pluribus, and bang-buck ratio positively drive PC1 scores. Everything else negatively drives PC1 scores. Interesting, there aren't any variables that are *not* correlated with this PC.





